---
title: "Practicum1"
author: "Matt Faucher, Kaelyn Jefferson"
date: "10/19/2021"
output:
  html_document: default
  pdf_document: default
  word_document: default
---
## Connect to the Database
```{r}
# 1. Library
library(RMySQL)
library(DBI)
# 2. Settings
db_user <- 'root'
db_password <- 'root'
db_name <- 'data'
db_host <- 'localhost'
db_port <- 3306
# 3. Read data from db
con <- dbConnect(MySQL(), user = db_user, password = db_password, dbname = db_name, host = db_host, port = db_port)
```
## Create Lookup table for Airports
```{sql connection=con}
CREATE TABLE IF NOT EXISTS airports (
  pid INTEGER NOT NULL UNIQUE AUTO_INCREMENT,
  code INTEGER NOT NULL UNIQUE,
  name VARCHAR(255) NOT NULL,
  city VARCHAR(255) DEFAULT 'N/A',
  state VARCHAR(255) NOT NULL DEFAULT 'N/A',
  country VARCHAR(255) NOT NULL,
  PRIMARY KEY (pid)
);
```
## Create Lookup table for Airlines
```{sql connection=con}
CREATE TABLE IF NOT EXISTS airlines (
  aid INTEGER NOT NULL UNIQUE AUTO_INCREMENT,
  code INTEGER NOT NULL UNIQUE,
  airline VARCHAR(255) NOT NULL,
  INDEX airline_idx (airline(255)),
  PRIMARY KEY (aid)
);
```
## Create Incidents Table
```{sql connection=con}
CREATE TABLE IF NOT EXISTS incidents (
  iid INTEGER NOT NULL,
  date DATE NOT NULL,
  depPort VARCHAR(255) NOT NULL DEFAULT 'unknown',
  arrPort VARCHAR(255) NOT NULL DEFAULT 'unknown',
  airline VARCHAR(255) NOT NULL DEFAULT 'unknown',
  aircraft VARCHAR(255) NOT NULL,
  flightPhase VARCHAR(255) NOT NULL,
  impact VARCHAR(255) NOT NULL DEFAULT 'none',
  PRIMARY KEY (iid)
);
```
## Add foreign key constraints
```{sql connection=con}
ALTER TABLE incidents
ADD CONSTRAINT FK_AIRLINE
FOREIGN KEY (airline) REFERENCES airlines(airline)
ON UPDATE CASCADE
ON DELETE CASCADE;
```
```{sql connection=con}
ALTER TABLE airlines
ADD CONSTRAINT
FOREIGN KEY (code) REFERENCES airports(code)
ON UPDATE CASCADE
ON DELETE CASCADE;
```

## Read Data from CSV (load)
```{r}
csvFile <- "BirdStrikesData.csv"
df.raw <- read.csv(file = csvFile, header = T, stringsAsFactors = F)
```
## Create dataframe for incidents table
```{r}
iid <- df.raw$Record.ID
date <- df.raw$FlightDate
date <- as.Date(date, format = "%m/%d/%Y")
depPort <- df.raw$Airport..Name
arrPort <- df.raw$Airport..Name
airline <- df.raw$Aircraft..Airline.Operator
aircraft <- df.raw$Aircraft..Make.Model
flightPhase <- df.raw$When..Phase.of.flight
# Harmonize flight phases
for (i in 1: length(flightPhase)) 
  if (flightPhase[i] == 'Take-off run'){
    flightPhase[i] <- 'takeoff'
  } else if (flightPhase[i] %in% c('Landing roll', 'Approach')){
    flightPhase[i] <- 'landing'
  } else if (flightPhase[i] %in% c('Climb', 'Descent')){
    flightPhase[i] <- 'inflight'
  } else flightPhase[i] <- 'unknown'

impact <- df.raw$Effect..Indicated.Damage
```
## Place all data into incidents table
```{r}
# Build the data frame
incidents <- data.frame(iid, date, depPort, arrPort, airline, aircraft, flightPhase, impact)
dbWriteTable(con, "incidents", incidents, overwrite = T, row.names = F)
```
## Get all data for airports
```{r}
pid <- c(1:25558)
code <- c(1:25558)
name <- df.raw$Airport..Name
city <- c()
for (i in 1:25558) {
  city[i] <- 'N/A'
}
state <- df.raw$Origin.State
country <- df.raw$Origin.State;
for (i in 1:length(country)) {
  if (country[i] != 'N/A') {
    country[i] <- 'USA'
  } else {
    country[i] <- 'N/A'
  }
}
```
## Build table for airports
```{r}
airports <- data.frame(pid, code, name, city, state, country)
dbWriteTable(con, "airports", airports, append = T, row.names = F)
```
## Create dataframe for airlines table & write table
```{r}
airline <- df.raw$Aircraft..Airline.Operator
# Build the data frame
aid <- c(1:25558)
code <- c(1:25558)
airlines <- data.frame(aid, code, airline)
dbWriteTable(con, "airlines", airlines, overwrite = T, row.names = F)
```
## Place defaults in for empty values in incidents table
```{sql connection=con}
UPDATE incidents
SET depPort = 'unknown', arrPort = 'unknown', airline = 'unknown'
WHERE depPort = '' AND arrPort = '' AND airline = '';
```
```{sql connection=con}
UPDATE airlines
SET airline = 'unknown'
WHERE airline = '';
```
## Data Populated in the Inicidents table
```{sql connection=con}
SELECT * FROM incidents LIMIT 5;
```
The incidents table (showing 5 entries), we decided to use the Record.Id as the
iid since this value was unique and acting as a primary key for records. The
departing and arriving airports were listed as the same airport since we lacked
data. Airlines that didn't have a value were modified to be "unknown". Flight
Phase has been harmonized to be one of 4 categories (inflight, takeoff, landing,
unknown). Takeoff consists of (Take-off run), inflight consists of ('Climb',
'Descent'), and landing consists of (Landing roll, Approach), and anything else
was categorized as Unknown.

## Data Populated in the Airports table
```{sql connection=con}
SELECT * FROM airports LIMIT 5;
```
The airports table (showing 5 entries), has the pid and code values. These values were synthetically filled with unique numbers 1:25558. For the city column, we added N/A values since we were not provided with data. For country, if there was no state then we considered the country to be unknown (outside the US).

## Data Populated in the Airlines table
```{sql connection=con}
SELECT * FROM airlines LIMIT 5;
```
The airlines table (showing 5 entries), we synthetically created the aid with values 1:25558. The code is the foreign key to the airports table (also synthetically created).

## Create a SQL query against your database to find the number of bird strike incidents for each airline arriving at LaGuardia airport during any phase of landing
```{sql connection=con}
SELECT airline AS Airline, COUNT(iid) AS Num_Incidents FROM incidents
WHERE arrPort LIKE 'LAGUARDIA NY'
AND flightPhase LIKE 'LANDING'
GROUP BY Airline
ORDER BY Num_Incidents DESC;
```
## Create a SQL query against your database to find the airport that had the most bird strike incidents (during any flight phase). Include all commercial airlines, i.e., no business, private, or military flights.
```{sql connection=con}
SELECT counts.airline as Commercial_Airline, MAX(counts.count) as Max_Incidents
FROM incidents as i,
	(SELECT airline, COUNT(iid) as count FROM incidents
    WHERE airline NOT LIKE '%business%' AND
    airline NOT LIKE '%jet%' AND
    airline NOT LIKE '%military%' AND
    airline NOT LIKE '%cargo%' AND
    airline NOT LIKE '%private%'AND
    airline NOT LIKE '%logistics%'AND
    airline NOT LIKE '%fedex%' AND
    airline NOT LIKE '%government%' AND
    airline NOT LIKE '%flight options%' AND
    airline NOT LIKE '%unknown%'
    GROUP BY airline) as counts
GROUP BY counts.airline
order by Max_Incidents DESC LIMIT 1;
```
In order to break down the airlines into groups so we could exclude everything that is not commercial, we added conditions to check for relevant keywords / substrings. The keywords we added filters for we know to not be commercial airlines based on the context of the words.

## Create a SQL query against your database to find the number of bird strike incidents by year.
```{r}
query <- "SELECT YEAR(date) AS Year, COUNT(iid) as Num_Incidents FROM incidents
WHERE date IS NOT NULL
GROUP BY Year"
incidentsPerYear <- dbGetQuery(con, query)
print(incidentsPerYear)
```

## Using the above data, build a line chart that visualizes the number of bird strikes incidents per year from 2005 to 2011. Adorn the graph with appropriate axis labels, titles, legend, data labels, etc.
```{r}
data <- incidentsPerYear[(incidentsPerYear$Year > 2004) & (incidentsPerYear$Year < 2012),]
data <- data[order(data$Year),]
x <- data$Year
y <- data$Num_Incidents
type = "b"
xlab = "Year"
ylab = "Num Incidents"
main = "Number of Bird Strike Incidents per Year"
xlim <- c(2005, 2011)
ylim <- c(1700, 3300)
plot(x = x, y = y, type = type, xlab = xlab, ylab = ylab, main = main, xlim = xlim, ylim = ylim)
```

## Create a stored procedure in MySQL that removes a bird strike incident from the database. You may decide what you need to pass to the stored procedure to remove a bird strike incident, e.g., departure airport, airlines, or some ID. Show that the deletion worked as expected.
```{sql connection=con}
DROP PROCEDURE IF EXISTS RemoveUnknownAirlines;
```
```{sql connection=con}
CREATE PROCEDURE RemoveUnknownAirlines()
BEGIN
	DELETE incidents, airlines
	FROM incidents, airlines
    WHERE incidents.airline = 'unknown' AND incidents.airline = airlines.airline;
END;
```
## Call Stored Procedure
```{sql connection=con}
CALL RemoveUnknownAirlines();
```

## Disconnect from the Databse
```{r}
dbDisconnect(con)
```